Supplementary Information for: Crosslinguistic Word Orders Enable an Efficient Tradeoff between Memory and Surprisal

### 2020-02-05 writeup/si.tex

\begin{center}
\includegraphics[width=0.5\textwidth]{../code/visualize_neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
\captionof{figure}{Histogram}\label{fig:hist-real}
\end{center}

@@

We formalize a language as a stationary stochastic process $\dots w_{-2} w_{-1} w_0 w_{1} w_{2} \dots$, extending indefinitely both into the past and into the future.
The symbols $w_i$ belong to a common set, representing the words of the language.\footnote{Could also be phonemes, sentences, ..., any other kind of unit.}
We denote the listener's memory state at time $t$, after hearing $w_{<t} = ... w_{t-2} w_{t-1}$ by $m_t$.
As described above, we assume
\begin{equation}\label{eq:listener-markov}
	p(m_{t+1}| (w_{t'})_{t' \in \mathbb{Z}}, m_t)   = p(m_{t+1} | m_t, w_{t})
\end{equation}
that is, $m_{t+1}$ contains no information about the utterances beyond what is contained in $m_t$ and $w_{t}$.
As a consequence, the listener has no knowledge of the speaker's state beyond the information provided in their prior communication.

@@

We first show a lemma:

\begin{lemma}
For any positive integer $t$, the following inequality holds:
\begin{equation}
H[w_t | m_t] \geq H[w_t|w_{1 \dots t-1}, m_1]
\end{equation}
\end{lemma}
\begin{proof}[Proof of the Lemma]
	By Bayes' Theorem
\begin{align*}
	p(w_t|m_0, m_1, w_{0\dots t-1}) &= \frac{p(m_1|m_0, w_{0\dots t})}{p(m_1|m_0, w_{0\dots t-1})} \cdot p(w_t|m_0, w_{0\dots t-1})
\end{align*}
By Equation~\ref{eq:listener-markov}, the quotient on the RHS is equal to $1$, so
\begin{align*}
p(w_t|m_0, m_1, w_{0\dots t-1}) = p(w_t|m_0, w_{0\dots t-1})
\end{align*}
So we have a Markov chain
\begin{equation}
(w_t) \rightarrow (m_0, w_{0 \dots t-1})   \rightarrow   (m_1, w_{1 \dots t-1})
\end{equation}
Thus, by the Data Processing Inequality,
\begin{equation}
H[w_t| w_{1 \dots t-1}, m_{1}] \geq H[w_t|w_{0 \dots t-1}, m_0]
\end{equation}
Finally, iteratively applying this inequality, we get:
\begin{align*}
H[w_t | m_t] \geq H[w_t| w_{t-1}, m_{t-1}] \geq H[w_t| w_{t-2, t-1}, m_{t-2}] \geq ... \geq H[w_t|w_{1 \dots t-1}, m_1]
\end{align*}
\end{proof}
Plugging this inequality into Equation~\ref{eq:byStation} above:
\begin{align*}
\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \sum_{t=1}^T ( \operatorname{H}[w_t|w_{1\dots t-1}, m_1] - \operatorname{H}[w_t | w_{1\dots t-1}, w_{\leq 0}]  )    \\
& = \frac{1}{T} \left(\operatorname{H}[w_{1\dots T} | m_1] - \operatorname{H}[w_{1\dots T} | w_{\leq 0}]\right)  \\
& = \frac{1}{T} \left(I[w_{1\dots T}|w_{\leq 0}] - I[w_{1\dots T}|m_1]\right) 
\end{align*}
The first term $I[w_{1\dots T}|w_{\leq 0}]$ can be rewritten in terms of $I_t$:
\begin{align*}
I[w_{1\dots T}|w_{\leq 0}] &= \sum_{i=1}^T \sum_{j=-1}^{-\infty} I[w_i, w_j | w_{j+1}...w_{i-1}] = \sum_{t=1}^T t I_t + T \sum_{t > T} I_t
\end{align*}
Therefore
\begin{align*}
\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - I[w_{1\dots T}|m_1]\right) 
\end{align*}
$I[w_{1\dots T}|m_1]$ is at most $\operatorname{H}[m_1]$, which is at most $\sum_{t=1}^T t I_t$ by assumption. Thus, the expression above is bounded by
\begin{align*}
\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - \sum_{t=1}^T t I_t\right) \\
&= \sum_{t > T} I_t
\end{align*}
Rearranging shows that the listener's surprisal is at least $\operatorname{H}[w_t|m_t] \geq \operatorname{H}[w_t | w_{<t}] + \sum_{t > T} I_t$, as claimed.
%\end{proof}

@@

%Bounding retrieval = bounding the precision and/or content of retrieved items. Explain more how this relates to McElree and ACT-R.

@@

\begin{proof}
The negative surprisal gap is
$\leq \frac{1}{T}(I[X_1\dots X_T|(M_0, R_1, ..., R_T)] - \sum_{t=1}^T t I_t - T \sum_{t>T} I_t) \leq  \frac{1}{T}(T\cdot H[R_t] - T \sum_{t>T} I_t) = H[R_t] - \sum_{t>T} I_t$

@@

>>>>

\begin{center}
\includegraphics[width=0.5\textwidth]{../code/analysis/visualize_neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
\captionof{figure}{Histogram}\label{fig:hist-real}
\end{center}

@@

We formalize a language as a stationary stochastic process $\dots w_{-2} w_{-1} w_0 w_{1} w_{2} \dots$, extending indefinitely both into the past and into the future.
The symbols $w_i$ belong to a common set, representing the words of the language.\footnote{Could also be phonemes, sentences, ..., any other kind of unit.}
We denote the listener's memory state at time $t$, after hearing $w_{<t} = ... w_{t-2} w_{t-1}$ by $m_t$.
As described above, we assume
\begin{equation}
	m_t = M(m_{t-1}, w_{t-1})
\end{equation}
\footnote{Alternatively we could admit nondeterministic memory encodings, and require
\begin{equation}\label{eq:listener-markov}
	p(m_{t+1}| (w_{t'})_{t' \in \mathbb{Z}}, m_t)   = p(m_{t+1} | m_t, w_{t})
\end{equation}
that is, $m_{t+1}$ contains no information about the utterances beyond what is contained in $m_t$ and $w_{t}$.}
As a consequence, the listener has no knowledge of the speaker's state beyond the information provided in their prior communication.

@@

%NOTE the proof could be made simpler by taking $m_t$ to be a deterministic function of $x_t$, $m_{t-1}$, rather than talking about conditional independence
%
%We first show a lemma:
%
%\begin{lemma}
%For any positive integer $t$, the following inequality holds:
%\begin{equation}
%H[w_t | m_t] \geq H[w_t|w_{1 \dots t-1}, m_1]
%\end{equation}
%\end{lemma}
%
%\begin{proof}[Proof of the Lemma]
Because $m_t$ is determined by $(w_{1 \dots t-1}, m_1)$:
\begin{equation}
	m_t = M(m_{t-1}, w_{t-1}) = M(M(m_{t-2}, w_{t-2}), w_{t-1}) = M(M(M(m_{t-3}, w_{t-3}), w_{t-2}), w_{t-1}) = \dots
\end{equation}
the Data Processing inequality entails the following inequality for every positive integer $t$:
\begin{equation}
	H[w_t | m_t] \geq H[w_t|w_{1 \dots t-1}, m_1]
\end{equation}

%\end{proof}

%\begin{proof}[Proof of the Lemma]
%	By Bayes' Theorem
%\begin{align*}
%	p(w_t|m_0, m_1, w_{0\dots t-1}) &= \frac{p(m_1|m_0, w_{0\dots t})}{p(m_1|m_0, w_{0\dots t-1})} \cdot p(w_t|m_0, w_{0\dots t-1})
%\end{align*}
%By Equation~\ref{eq:listener-markov}, the quotient on the RHS is equal to $1$, so
%\begin{align*}
%p(w_t|m_0, m_1, w_{0\dots t-1}) = p(w_t|m_0, w_{0\dots t-1})
%\end{align*}
%So we have a Markov chain
%\begin{equation}
%(w_t) \rightarrow (m_0, w_{0 \dots t-1})   \rightarrow   (m_1, w_{1 \dots t-1})
%\end{equation}
%Thus, by the Data Processing Inequality,
%\begin{equation}
%H[w_t| w_{1 \dots t-1}, m_{1}] \geq H[w_t|w_{0 \dots t-1}, m_0]
%\end{equation}
%Finally, iteratively applying this inequality, we get:
%\begin{align*}
%H[w_t | m_t] \geq H[w_t| w_{t-1}, m_{t-1}] \geq H[w_t| w_{t-2, t-1}, m_{t-2}] \geq ... \geq H[w_t|w_{1 \dots t-1}, m_1]
%\end{align*}
%\end{proof}
Plugging this inequality into Equation~\ref{eq:byStation} above:
\begin{align*}
\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \sum_{t=1}^T ( \operatorname{H}[w_t|w_{1\dots t-1}, m_1] - \operatorname{H}[w_t | w_{1\dots t-1}, w_{\leq 0}]  )    \\
& = \frac{1}{T} \left(\operatorname{H}[w_{1\dots T} | m_1] - \operatorname{H}[w_{1\dots T} | w_{\leq 0}]\right)  \\
& = \frac{1}{T} \left(I[w_{1\dots T}, w_{\leq 0}] - I[w_{1\dots T}, m_1]\right) 
\end{align*}
The first term $I[w_{1\dots T}, w_{\leq 0}]$ can be rewritten in terms of $I_t$:
\begin{align*}
I[w_{1\dots T}, w_{\leq 0}] &= \sum_{i=1}^T \sum_{j=-1}^{-\infty} I[w_i, w_j | w_{j+1}...w_{i-1}] = \sum_{t=1}^T t I_t + T \sum_{t > T} I_t
\end{align*}
Therefore
\begin{align*}
\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - I[w_{1\dots T}, m_1]\right) 
\end{align*}
$I[w_{1\dots T}|m_1]$ is at most $\operatorname{H}[m_1]$, which is at most $\sum_{t=1}^T t I_t$ by assumption. Thus, the expression above is bounded by
\begin{align*}
\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - \sum_{t=1}^T t I_t\right) \\
&= \sum_{t > T} I_t
\end{align*}
Rearranging shows that the listener's surprisal is at least $\operatorname{H}[w_t|m_t] \geq \operatorname{H}[w_t | w_{<t}] + \sum_{t > T} I_t$, as claimed.
%\end{proof}

@@

Bounding retrieval = bounding the precision and/or content of retrieved items. Explain more how this relates to McElree and ACT-R.

In ACT-R, each retrieval operation takes time. If each chunk has a bounded amount of information, then this corresponds to a bottleneck in $H[r_t|m_t]$

Formally:

$m_t = f(x_t, m_{t-1}, r_t)$

$r_t = g(m_{t-1}, s_{t-1})$

$s_t = h(m_{t-1}, s_{t-1})$

@@

\begin{proof}
	As before using the Data Processing inequality:
\begin{align*}
	-(\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]) \leq & \frac{1}{T}\left(I[X_1\dots X_T, (M_0, R_1, ..., R_T)] - I[X_1\dots X_T, X_{<1}]\right) \\
= & \frac{1}{T}\left(I[X_1\dots X_T, (M_0, R_1, ..., R_T)] - \sum_{t=1}^T t I_t - T \sum_{t>T} I_t\right) \\
	= & \frac{1}{T}\left(I[X_{1\dots T}, M_0] + I[X_{1\dots T}, R_1|M_0] + \dots + I[X_{1\dots T}, R_T|M_0, R_{1 \dots T-1}] - \sum_{t=1}^T t I_t - T \sum_{t>T} I_t\right) \\
	\leq & \frac{1}{T}\left(H[M_0] + H[R_1] + \dots + H[R_{T-1}] - \sum_{t=1}^T t I_t - T \sum_{t>T} I_t\right) \\
	\leq &  \frac{1}{T}(T\cdot H[R_t] - T \sum_{t>T} I_t) \\
	= & H[R_t] - \sum_{t>T} I_t
\end{align*}

@@

\subsection{Results for Language Production}

Here we show results linking memory and locality in production.

First, we consider a setting in which a speaker produces sentences with bounded memory, and analyze the deviation of the produced distribution from the actual distribution of the language.

The first setting does not account for the fact that language is produced aiming for some communicative goal.
We therefore now assume that the speaker has a communicative goal $G$ in mind.
This goal $G$ stays constant during production process for a sentence, and we count how much memory is needed in addition to the goal $G$.
We assume that there is a distribution of sentences expressing goals $G$:
\begin{equation}
	P(sentence|G)
\end{equation}
and assume that the speaker aims to match this distribution
\begin{equation}
\mathbb{E}_G[D_{KL}((language|G)||(produced|G))]
\end{equation}
We can analyze this model by adding conditioning w.r.t. $G$ throughout the analysis of the previous case.
Specifically, we need $I_t^G := I[X_t, X_0|X_1, \dots, X_{t-1}, G]$.

Take $I_t$ conditioned on $G$: only count statistical dependencies to the degree that they are not redundant with the goal
